---
title: "Exercises Day 2"
author: "Rachel Rosenberg"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{asis}
### Answer
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{asis}
### Answer
```

```{r}
for (i in c(2002, 2012)){ # added for loop to go through 2002 and 2012
if(any(gapminder$year == i)){
   print(paste("Record(s) for the year",i,"found."))
} else {
  print(paste("No records for year",i))
}
} 
```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{asis}
### Answer
```

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```


## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{asis}
### Answer

Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/
```

```{r}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{asis}
### Answer

```

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r}
relation_gdp_le <- function(df, year){
  dat <- df[df[["year"]]==year, ]
  x <- log(dat$gdpPercap)
  y <- log(dat$lifeExp)
  fit <- lm(y ~ x)
  plot(y ~ x, main = "Relationship between per capita GDP and Life Expectancy",
       ylab = "Log of Life Exp", xlab = "Log of Per Capita GDP")
}
relation_gdp_le(gapminder, 2002)
```


2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.