---
title: "R Day 2"
author: "Rachel Rosenberg"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data","gapminder5.csv") # adds /data/ and /gapminder5.csv to the folder path

gapminder <- read.csv(here::here("data","gapminder5.csv")) # reads the CSV file - you do this b/c everyone's file path may be different. When you use here::here, everyone's unique "home" part of the directory is auto populated.

```

## Converting factors to characters

```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)

```

## Find natural log of GDP per capita and population with a for loop

```{r}
obs <- 1:nrow(gapminder)

for (i in obs){
  gapminder[i, "log_gdpPercap"] <- log(gapminder[i, "gdpPercap"])
  gapminder[i, "log_pop"] <- log(gapminder[i, "pop"])
}
head(gapminder)

## double check that it worked
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap == gapminder$log_gdpPercap)
```

## Find if life expectancy has increased over time and find the mean life expectancy per year

```{r}

years <- unique(gapminder$year)

for (i in years){
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], na.rm = TRUE) # subset the data frame, select the lifeExp variable, and find the mean in the year i.
  print(paste0(i, ": ", mean_le)) # paste command pastes strings together - in this case it will tell you the mean in each year. If you didn't specifically print, it won't display. Print and paste are often found together
  
}
```

## do the same thing for continents. which continent has the highest mean life expectancy?

```{r}
conts <- unique(gapminder$continent)

for (i in conts){
  mean_lifeExp <- mean(gapminder$lifeExp[gapminder$continent == i], na.rm = TRUE)
  print(paste0(i, ": ", mean_lifeExp))
}
```

## Practice nested for loops. What is mean_le for each country for each year?

```{r}
for (i in conts){
  print(paste0("Continent: ", i))
  for(j in years){
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j], na.rm=)
    print(paste0(j, ": ", mean_le))
  }
}
```

## Has the gap in life expectancy between countries on different continents narrowed over time?

```{r}
# Use standard deviation - if stdev is getting smaller, then there's less variance
# iterate over continents and time

for (i in conts){
  print(paste0("Continent: ", i))
  for (j in years){
    sd_le <- sd(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j], na.rm = T)
    print(paste0(j, ": ", sd_le))
  }
}
```

## Using the apply family of functions
## apply(matrix, 1 = row or 2 = col, function)

```{r}
vars <- gapminder[,c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean, na.rm = T)

# try lapply
lapply(gapminder, mean, na.rm = T)

# use apply with anonymous functions
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year == x]))
```

## While loops practice

```{r}
i <- 1952
while(i < 1987){
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le))
  i <- i + 5
}
```

## What is the standard deviation for life expectancy for each year between 1987 and 2002 (inclusive)?

```{r}
i <- 1987
while(i <= 2002){
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i], na.rm = T)
  print(paste0(i, ": ", sd_le))
  i <- i + 5
}
# hit esc if you accidentally create an infinite loop
```

## Practice with if statements 

```{r}
set.seed(10)
random_year <- sample(years, 1)
if (random_year < 1977){
  print(random_year)
} else {
  print("sorry, your number is not less than 1977.")
}
```

## Putting for and if/else together

```{r}
# Which continents have a mean life expectancy greater than 70 years?
threshold <- 70

for (i in unique(gapminder$continent)){
  tmp <- mean(gapminder$lifeExp[gapminder$continent == i])
  if (tmp < threshold){
  print(paste("Mean Life Expectancy in ", i, "is less than ", threshold))
  } else {
    print(paste("Mean Life Expectancy in ", i, "is greater than ", threshold))
  }
}

# Write a for loop that reports the mean population for years greater than or equal to 1987. Make sure the loop prints a message if the condition is not met!

i <- 1987

for (i in years){
  mpop <- mean(gapminder$pop[gapminder$year == i], na.rm = T)
  if (i >= 1987){
    print(paste("The population in ", i, "was ", mpop))
  } else{
    print(paste("Sorry, that year is not in the range"))
  }
  i <- i + 5
}

```

## Writing functions
# if you have to copy and paste three times, it's time to write a function
```{r}
# print the value of a selected variable in the gapminder dataset
get_values <-
  function(df, variable = "continent"){
    vals <- unique(df[[variable]]) # get the unique values of that variable in my df. Use double bracket notation because it will just index the column and all of the rows. Use this instead of $ b/c the $ can't deal with a character string, which the variable variable is.
    print(paste0(variable, ": ", vals))
  }

# print the mean and stdev for life exp for a given country
report_mean_sd <-
  function(df, variable, country){
    var <- gapminder[[variable]][gapminder$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country: ", country, 
        "\nMean Life Expectancy: ", m_le, 
        "\nSD Life Expectancy", sd_le)
  }
```

## Write a function that reports the mean, median, minimum, and maximum for life expectancy for a continent in gapminder
## Hint: min, max

```{r}
metric_report <-
  function(df, continent = "Asia"){
    var <- gapminder$lifeExp[gapminder$continent == continent]
    mean_le <- mean(var)
    median_le <- median(var)
    min_le <- min(var)
    max_le <- max(var)
    cat("Continent: ", continent,
        "\nMean: ", mean_le,
        "\nMedian: ", median_le,
        "\nMin: ", min_le,
        "\nMax: ", max_le)
  }
```

## Combine loops and functions

```{r}
viz_lm <-
  function(df, dv, iv, year){
    dat <- df[df[["year"]]== year, ] #subset by year
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x) #do a simple linear regression of y on x
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = "blue") # add a line that is the predicted value for the dependent variable based on the linear regression
  }

viz_lm(gapminder, "lifeExp", "gdpPercap", 1977)

# or can use for loops to iterate over years

for (i in years){
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```

